Finch: Sparse and Structured Array Programming with Control Flow

### 2024-03-22 sections/intro.tex
Update on Overleaf.

%Bring in structure
% - old arrays are Dense, rectilinear, integer grids
% - The world is not like that
% - Sparsity, runs of repeated values, symmetry
% - Lots of citations, perhaps a few figures from my powerpoint
%Sparse arrays
% - Networks (graphblas, graph frameworks)
% - Simulations (BLAS, symblas, banded blas, etc)
% - Databases (database engines)
%Run-length encoding
%Symmetry
%Bands
%Padding
%Blocks
\help{This section needs citations, please help expand it and make it more
convincing. A figure could be nice.} 
\begin{figure}
	\includegraphics[width=\linewidth]{example_structures.png}
    \caption{some example structures that might be nice to put into a figure}
\end{figure}
Our world is full of structured arrays.
%

Sparse arrays (which store only nonzero elements) describe networks and
simulations.
%
Run-length encoding describes images and masks, geometry, and
databases (such as a list of transactions with the date field all the same).
%
Symmetry, bands, padding, and blocks arise due to modeling choices in scientific computing as well as in intermediate structures in many linear solvers.
%
Operators, such as convolution, can be expressed as structured arrays.
%
For example, a convolution with a filter can be expressed as a matrix multiplication
with the toeplitz matrix of all the circular shifts of the filter~\cite{sze2017efficient}.
%
\help{This section needs citations}
\textbf{Currently, support for structured data is fragmented and incomplete}.
%
Experts must hand write variations of even the simplest kernels, like matrix
multiply, for each data structure and architecture to get performance.
%
Implementations must choose a small set of features to support well, resulting
in a compromise between \textbf{program flexibility} and \textbf{datastructure
flexibility}.
%
Hand-written solutions are collected in diverse libraries like
MKL, OpenCV, or SciPy. 
%
However, libraries will only ever support a subset of
programs on a subset of datastructure combinations. Even the most advanced
libraries, such as the GraphBLAS, which support a wide variety of sparse
operations over various semirings always lack support for other features, such
as tensors, fused outputs, or runs of repeated values. 
%
While dense array
compilers support an enormous variety of program constructs like early break and
multiple left hand sides, they only support dense arrays.  
%
Special-purpose
compilers like TACO or CoRa which support a select subset of structured data
structures (only sparse, or only ragged arrays) must compromise by greatly
constraining the classes of programs which they support, such as tensor
contractions \cite{kjolstad_tensor_2019, fegade_cora_2022}. 
%
This trade-off is
visualized in figure \ref{fig:datavsprogram}.

>>>>

%Bring in structure
% - old arrays are Dense, rectilinear, integer grids
% - The world is not like that
% - Sparsity, runs of repeated values, symmetry
% - Lots of citations, perhaps a few figures from my powerpoint
%Sparse arrays
% - Networks (graphblas, graph frameworks)
% - Simulations (BLAS, symblas, banded blas, etc)
% - Databases (database engines)
%Run-length encoding
%Symmetry
%Bands
%Padding
%Blocks
\help{This section needs citations, please help expand it and make it more
convincing. A figure could be nice.} 
\teo{I Added citations/heft but we maybe need to improve the figure?}
\begin{figure}
	\includegraphics[width=\linewidth]{example_structures.png}
    \caption{some example structures that might be nice to put into a figure}
\end{figure}
Our world is full of structured arrays.
%
Sparse arrays (which store only nonzero elements) describe networks, databases, and simulations~\cite{abhyankarpetsc, bell2007lessons, mcauley2013hidden, balay2020petsc}.
%
Run-length encoding describes images and masks, geometry, and
databases (such as a list of transactions with the date field all the same)~\cite{shi2020column,golomb1966run}.
%
Symmetry, bands, padding, and blocks arise due to modeling choices in scientific computing (e.g. higher order FEMs) as well as in intermediate structures in many linear solvers (e.g. GMRES)~\cite{ded, saad2003iterative, o2009scientific}.
%
In the context of machine learning, combinations of sparse and blocked matrices are increasingly under consideration~\cite{dao2022monarch}.
%
Operators, such as convolution, can be expressed as structured arrays.
%
For example, a convolution with a filter can be expressed as a matrix multiplication
with the toeplitz matrix of all the circular shifts of the filter~\cite{sze2017efficient}.

%
\help{This section needs citations}
\textbf{Currently, support for structured data is fragmented and incomplete}.
%
Experts must hand write variations of even the simplest kernels, like matrix
multiply, for each data structure/data set and architecture to get performance.
%
Implementations must choose a small set of features to support well, resulting
in a compromise between \textbf{program flexibility} and \textbf{data structure
flexibility}.
%
Hand-written solutions are collected in diverse libraries like
MKL, OpenCV, LAPACK or SciPy~\cite{ bradski2000opencv, anderson1999lapack, virtanen2020scipy, psarras2022linear}. 
%
However, libraries will only ever support a subset of
programs on a subset of data structure combinations.
%
Even the most advanced
libraries, such as the GraphBLAS, which support a wide variety of sparse
operations over various semi-rings always lack support for other features, such
as tensors, fused outputs, or runs of repeated values~\cite{bulucc2017design, mattson2019lagraph}.
%
While dense array
compilers support an enormous variety of program constructs like early break and
multiple left hand sides, they only support dense arrays~\cite{ragan-kelley_halide_2013,grosser2012polly}.  
%
Special-purpose
compilers like TACO, Taichi, StructTensor, or CoRa which support a select subset of structured data
structures (only sparse, or only ragged arrays) must compromise by greatly
constraining the classes of programs which they support, such as tensor
contractions \cite{kjolstad_tensor_2019, hu_taichi_2019, ghorbani2023compiling, fegade_cora_2022}. 
%
This trade-off is
visualized in figure \ref{fig:datavsprogram}.
